\section{Introduction}

Stainless \cite{HamzaETAL19SystemFR} 
is a tool for verifying and finding errors in
programs written in a subset of the Scala~\cite{OderskyETAL20ProgrammingScala} programming language.
Stainless is open source (distributed under Apache license) and hosted on GitHub at:
\begin{center}
  \url{https://github.com/epfl-lara/stainless/} \\
  \url{https://epfl-lara.github.io/stainless/}
\end{center}
Stainless (and its predecessor, Leon) have
been developed primarily in the EPFL's Laboratory for
Automated Reasoning and Analysis in the period from
  2011-2021, see, in particular \cite{HamzaETAL19SystemFR, voirol2019} as well as
  \cite{schmid2021proving,MadhavanKuncak17Memoization,blanc2017,
    VoirolETAL15CounterExampleCompleteVerificationHigherOrderFunctions,
    BlancKuncak15SoundReasoningIntegralDataTypes, Kuncak15DevelopingVerifiedSoftwareUsingLeonNFM,
    KoukoutosKuncak14CheckingDataStructurePropertiesOrdersMagnitudeFaster,
    BlancETAL13VerificationTranslationRecursiveFunctions, KoeksalETAL12ConstraintsControl,
    SuterETAL11SatisfiabilityModuloRecursivePrograms,SuterETAL10DecisionProceduresforAlgebraicDataTypesAbstractions}.
The core specification and implementation
language of Stainless are typed recursive higher-order
functional Scala programs. It also supports certain
imperative programs \cite{schmid2021proving,blanc2017}. Stainless can verify that functions
are correct for all inputs with respect to provided
preconditions and postconditions, it can prove that
functions terminate (with optionally provided termination
measure functions), and it can also provide
counter-examples to safety properties.

Stainless can be used to write programs that are directly
executable and proven correct. In particular, because it
uses Scala's syntax and type system, users can execute
Stainless programs using the standard Scala compiler
(version 2.12.13 at the time of writing). In addition, there are passes
that eliminate non-executable (ghost) code from source to make sure
that it does not result in run-time overhead after compilation. For
programs that adhere to certain discipline the ``genc'' option of Stainless can
be used to generate C source code that compiles with common
compilers such as gcc.

\subsection{Outline}
  
In this tutorial, we show examples
demonstrating how to use Stainless to develop verified
models and programs. We will mostly use basic notation
for functional programming, which we will introduce along the way.
We will use Stainless version 0.9 or later.

In addition to basic introduction, we will suggest
strategies for specifying programs and helping Stainless
prove them correct. An example is using lemmas and proving
them by induction expressed through terminating recursion.

To help users be more effective when using Stainless, we
also outline key mechanisms that Stainless uses
in proof and counterexample search: encoding into
functional programs, function unfolding, and using rich
theories of satisfiability modulo theory solvers Z3 and
CVC4.
