\documentclass{fmcad}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{sansmath}
\usepackage{url}
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
%\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}

\usepackage[final]{listings}
\input{lstsettings}

\begin{document}

\title{Stainless Verification System Tutorial}
\author{
  \IEEEauthorblockN{Viktor Kun\v{c}ak}
  \IEEEauthorblockA{\textit{LARA Research Group} \\
    \textit{School of Computer and Communication Sciences} \\
    \textit{EPFL}\\
    Lausanne, Switzerland \\
    viktor.kuncak@epfl.ch}
\and
  \IEEEauthorblockN{Jad Hamza}
  \IEEEauthorblockA{\textit{LARA Research Group} \\
    \textit{School of Computer and Communication Sciences} \\
    \textit{EPFL}\\
    Lausanne, Switzerland \\
    jad.hamza@epfl.ch}
}

\maketitle

\begin{abstract}
  Stainless ( https://stainless.epfl.ch ) is an open-source tool for
  verifying and finding errors in programs written in the
  Scala programming language. This tutorial will not assume
  any knowledge of Scala. It aims to get first-time users
  started with verification tasks by introducing the
  language, providing modelling and verification tips, and
  giving a glimpse of the tool's inner workings (encoding
  into functional programs, function unfolding, and using
  theories of satisfiability modulo theory solvers Z3 and
  CVC4).

  Stainless (and its predecessor, Leon) has been developed
  primarily in the EPFL's Laboratory for Automated Reasoning
  and Analysis in the period from 2011-2021. Its core
  specification and implementation language are typed
  recursive higher-order functional programs (imperative
  programs are also supported by automated translation to
  their functional semantics).  Stainless can verify that
  functions are correct for all inputs with respect to
  provided preconditions and postconditions, it can prove
  that functions terminate (with optionally provided
  termination measure functions), and it can provide
  counter-examples to safety properties.  Stainless enables
  users to write code that is both executed and verified
  using the same source files. Users can compile programs
  using the Scala compiler and run them on the JVM. For
  programs that adhere to certain discipline, users can
  generate source code in a small fragment of C and then
  use standard C compilers.
\end{abstract}

\begin{IEEEkeywords}
  verification, formal methods, proof, counter-example, model checking,
  Scala, functional programming, satisfiability modulo theories
\end{IEEEkeywords}

\input{intro}
\input{start}
\input{functional}
\input{queue}
\input{proofs}
\input{digits}
\input{imperative}
\input{design}
  
   \bigskip
   \noindent 
  {\textbf{Acknowledgements.}}\ 
   Research on Stainless has been funded in part
   by (i) the Swiss Science Foundation grants 200021\_132176, 200020\_138204, 200020\_146649, 200021\_144503, 200020\_159949,
   and 200021\_175676.   
   (ii) European Research Council (ERC) Starting Grant PE6-306484-IMPRO, 
   (iii) The Swiss State Secretariat for Education,
   Research and Innovation, Swiss Space Office grant ``Embedded Flight Software Verification--ESOVER'' and
   (iv) the envelope budget for the LARA group from the EPFL School of Computer and Communication Sciences.

   Stainless and Inox were created from parts of Leon code by Nicolas Voirol.
   In addition to Nicolas and the two authors of this tutorial, contributors to Stainless and Inox include:
   Roman Ruetschi, Georg Stefan Schmid, Marco Antognini, Ravichandhran Madhavan, Etienne Kneuss,    
   Lars Hupel, Emmanouil Koukoutos, Philippe Suter, Roman Edelmann, Utkarsh Upadhyay, Ivan Kuraj, Sandro Stucki,
   Ruzica Piskac, Tihomir Gvero,  Czipó Bence, Sumith Kulal, Lucien Iseli,   
   Regis Blanc, Iulian Dragos, Dragana Milovan\v{c}evi\'c, Antoine Brunner, Mirco Dotta,   
   Yann Bolliger, Rodrigo Raya, Samuel Gruetter, Mika\"el Mayer, Guillaume Massé.
   Romain Jufer worked with Jad Hamza on a fork for smart contract verification and Solidity code generation,
   Romain Edelmann and Rodrigo Raya developed an interactive proof assistant concept based on Inox.
   Regis Blanc developed a Scala library for
   input and output of SMT-LIB files. ScalaZ3 interface to the Z3 dynamically
   linked library additionally received contributions from Ali Sinan K\"oksal
   and Thorsten Tarrach.
   Contributors to Stainless Bolts case studies include additionally Samuel Chassot and Cl\'ement Burgelin. We thank
   users of Stainless from Ateleris GmbH including Simon Felix, Filip Schramka, and Ivo Nussbaumer. We also thank
   MSc students at EPFL taking the Formal Verification course, completing interesting case studies and identifying bugs in the system.
\bibliographystyle{IEEEtran}
\bibliography{main}

\end{document}
